A Bidirected String Graph Model for Genome Assembly

### 2012-12-08 presentation/paper.tex
paper: graph building

\newcommand{\LengthVar}{L}
\newcommand\Base[1]{{\tt #1}}
\newcommand{\ReadArrowType}{latex}
\newcommand{\Figure}[1]{Figure \ref{#1}}

@@

@@

@@

Consider the the string graph fragment shown in \Figure{fig:bidiedge_interp_a}.
When we walk from $f \to g$, we are using both $f$ and $g$ in forward
orientation, based on the arrow heads.  But what this really means is that we
are walking from the {\em end} of read $f$ to the {\em end} of read $g$.  In
other words, traversing the edge in that direction is associated with the
addition of the sequence after the overlapped region in $g$, up to the end of
the $g$.

@@

\begin{figure}
	\centering
	\begin{subfigure}{0.48\textwidth}
		\begin{tikzpicture}[>=triangle 45,scale=1.5]
			\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
			\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
			\draw[>->,style=very thick] (f) edge (g);
			\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
			\draw (1.00, -0.1) node[anchor=north] {
					\rotatebox{180}{\small \tt
					TGTGTAATC}};
			\draw[->] (0.5, 0.5) ..
				controls (0.7, 0.7) and (1.3, 0.7) ..
				(1.5, 0.5);
			\draw[->] (1.5, -0.5) ..
				controls (1.3, -0.7) and (0.7, -0.7) ..
				(0.5, -0.5);
		\end{tikzpicture}
		\caption{bidirected string graph fragment}
		\label{fig:bidiedge_interp_a}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
			\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
					-> (4, 0) node[anchor=west] {$f.E$};
			\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
					-> (6.09, -0.9) node[anchor=west] {$g.E$};

			\draw (0, -0.1) node[anchor=south west]
				{\hspace*{-1.4mm}{\small \tt
					%GATTACACAGTGTAGTT
					GATTACACA\hspace*{0.1cm}$a_1 \dots a_n$
				}};

@@

			\draw (2.09, -1.4) node[anchor=north] {$o.g_{begin}$}
					-- (2.09, 0.6) node[anchor=south] {$o.f_{begin}$};
			\draw (3.99, -1.4) node[anchor=north] {$o.g_{end}$}
					-- (3.99, 0.6) node[anchor=south] {$o.f_{end}$};
		\end{tikzpicture}
		\caption{The corresponding overlap (labeled)}
		\label{fig:bidiedge_interp_b}
	\end{subfigure}

	\begin{subfigure}{0.5\textwidth}
		\begin{tabular}{c|c}
			walk & interpretation \\ \hline
			$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
			length}$(g)]$ \\
			$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
		\end{tabular}
		\caption{Interpretation of each walk on the graph fragment}
	\end{subfigure}
	\caption{Full interpetation of an edge in the bidirected string graph}
	\label{bidiedge_interp}
\end{figure}

\section{Removing contained reads}

There is yet another flaw in the way that the construction of the graph has been
presented that needs to be addressed.  Earlier, the definition of a
\KeyTerm{contained read} was provided.

>>>>

\newcommand{\LengthVar}{L}
\newcommand\Base[1]{{\tt #1}}
\newcommand{\ReadArrowType}{latex}
\newcommand{\Figure}[1]{Figure \ref{#1}}
\newcommand{\Subfigure}[1]{Figure \ref{#1}}
\newcommand{\Section}[1]{Section \ref{#1}}
\newcommand{\Subsection}[1]{Subsection \ref{#1}}

@@

\label{sec:build_graph}

@@

\label{subsec:graph_overview}

@@

Consider the the string graph fragment shown in
\Subfigure{subfig:bidiedge_interp_a}.  When we walk from $f \to g$, we are using
both $f$ and $g$ in forward orientation, based on the arrow heads.  But what
this really means is that we are walking from the {\em end} of read $f$ to the
{\em end} of read $g$.  In other words, traversing the edge in that direction is
associated with the addition of the sequence after the overlapped region in $g$,
up to the end of the $g$.

@@

\begin{figure}
	\begin{center}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
				\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
				\draw[>->,style=very thick] (f) edge (g);
				\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.00, -0.1) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
				\draw[->] (0.5, 0.5) ..
					controls (0.7, 0.7) and (1.3, 0.7) ..
					(1.5, 0.5);
				\draw[->] (1.5, -0.5) ..
					controls (1.3, -0.7) and (0.7, -0.7) ..
					(0.5, -0.5);
			\end{tikzpicture}
			\caption{A bidirected string graph fragment}
			\label{subfig:bidiedge_interp_a}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
				\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
						-> (4, 0) node[anchor=west] {$f.E$};
				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
						-> (6.09, -0.9) node[anchor=west] {$g.E$};

				\draw (0, -0.1) node[anchor=south west]
					{\hspace*{-1.4mm}{\small \tt
						%GATTACACAGTGTAGTT
						GATTACACA\hspace*{0.26cm}$a_1 \dots a_n$
					}};

@@

				\draw (2.09, -1.4) node[anchor=north] {$o.g_{begin}$}
						-- (2.09, 0.6) node[anchor=south] {$o.f_{begin}$};
				\draw (3.99, -1.4) node[anchor=north] {$o.g_{end}$}
						-- (3.99, 0.6) node[anchor=south] {$o.f_{end}$};
			\end{tikzpicture}
			\caption{The corresponding overlap (labeled)}
			\label{subfig:bidiedge_interp_b}
		\end{subfigure}
	\end{center}

	\begin{subfigure}{1.00\textwidth}
		\begin{center}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=blue!20,draw=black] (fE) at (0, 0) {f.E};
				\node[circle,fill=blue!20,draw=black] (gE) at (2.0, 0) {g.E};

				\node[circle,fill=blue!20,draw=black] (fB) at (0, -1) {f.B};
				\node[circle,fill=blue!20,draw=black] (gB) at (2.0, -1) {g.B};

				\draw[->,style=very thick] (fE) edge (gE);
				\draw[->,style=very thick] (gB) edge (fB);
				\draw (0.90, 0.01) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.10, -1.01) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
			\end{tikzpicture}
		\end{center}
		\caption{The corresponding directed string graph fragment}
		\label{subfig:bidiedge_interp_a}
	\end{subfigure}

	\vspace{1.0cm}

	\begin{subfigure}{1.00\textwidth}
		\begin{center}
			\begin{tabular}{c|c}
				walk & interpretation \\ \hline
				$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
				length}$(g)]$ \\
				$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
			\end{tabular}
		\end{center}
		\caption{Interpretation of each walk on the bidirected graph fragment}
	\end{subfigure}
	\caption{Full interpetation of an edge in a bidirected string graph}
	\label{fig:bidiedge_interp}
\end{figure}

\subsection{An equivalent directed graph}

\label{subsec:directed}

In \Subsection{subsec:graph_overview}, the possibility of using a directed graph
to model the assembly of the reads was rejected because it did not take into
account the double-strandedness of DNA.  However, \Figure{fig:bidiedge_interp}
suggests an alternate construction of the graph that does indeed use directed
edges.  Instead of having each vertex be a read, have each vertex be an end of a
read.  The advantage to this approach is that the problem can be expressed using
a directed graph instead a bidirected graph, which can be helpful for the
implementation.  The disadvantage is that two vertices are required for each
read, they may be widely separated in the graph.

\subsection{Removing contained reads}

\label{subsec:removing_contained_reads}

Earlier, the definition of a \KeyTerm{contained read} was provided.  Contained
reads need not be included in the graph because there will always be a walk that
spells the sequence of each contained read.  In other words, contained reads
provide redundant information because the sequence of any given contained read
is already spelled out by an uncontained read.

Therefore, prior to mapping the reads to vertices and overlaps to edges, we can
examine the overlaps to identify contained reads, and include in the initial
construction of the bidirected string graph neither the contained reads nor any
overlaps involving a contained read.

The contained reads cannot, however be fully discarded, as they are needed for a
later step (see \Section{sec:mapping_contained_reads}).

\subsection{Implementation}

As described in \Subsection{subsec:removing_contained_reads}, before we even
build the graph, contained reads need to be removed.  This is implemented by the
program \ProgramName{remove-contained-reads}.  It takes as input two files: the
set of reads and the set of overlaps that were computed from them, and produces
as output three files: the set of uncontained reads, the set of overlaps
excluding those that involve contained reads, and an array that maps from the
original read indices to the new read indices.

Building the graph is implemented by either the program
\ProgramName{build-bidirected-string-graph} or the program
\ProgramName{build-directed-string-graph}, the latter of which builds the
directed equivalent to the bidirected graph described in
\Subsection{subsec:directed}.  Both programs take as input two files: the set of
uncontained reads and their overlaps, and produce as output the bidirected or
directed graph, serialized to a file.  As should be expected from what was
described earlier, \ProgramName{build-bidirected-string-graph} will add one
vertex for each read and add a bidirected edge for each overlap, while
\ProgramName{build-directed-string-graph} will add two vertices for each read
and add two directed edges for each overlap.

The string graphs are stored in a binary format.  An additional program,
\ProgramName{print-string-graph}, was implemented that can print a string graph
in plain-text format, or print a string graph in the DOT language for input to
the graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print
some other information about a string graph.

Furthermore, programs \ProgramName{bidigraph-to-digraph} and
\ProgramName{digraph-to-bidigraph} were implemented to convert between the
bidirected and directed representations of the string graph.

A program \ProgramName{extract-edge-seqs} was implemented to extract the
sequences that the edges are labeled with, as this can be done to produce a
preliminary assembly (especially after the graph simplifications that are about
to be described).

\section{Transitive reduction}

Before trying to find walks through the string graph, it is very helpful to
simplify the graph is much as

\section{Mapping contained reads}
\label{sec:mapping_contained_reads}
